Typing a linear \picalc

### 2020-04-13 paper/paper.tex
Add weakening and strengthening theorems

In the \lambdacalc{} $\beta$-reduction operates on syntactically adjacent terms.
In the \picalc{} however, the syntax introduces unnecessary distinctions
(e.g. semantically parallel composition is defined modulo associativity and commutativity).
There are several ways around this, a structural congruence relation being one of the historical ones.
(Others include labeled transition systems and higher inductive types.)

@@

Structural congruence is a congruent equivalence relation on processes.
Any two structurally congruent processes are strongly bisimilar: they are can follow each other's reduction steps \cite{}.
Figure \ref{fig:struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_0 \; Q$ witnesses that index $0$ does not appear nor in the inputs nor in the outputs of process $Q$ --- it does so by traversing $Q$.
The function $lower \; 0 \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $0$.
Finally, $swap \; 0 \; P$ traverses $P$ (scoped under $\suc \suc n$) and swaps variable references $0$ and $\suc 0$.

@@

\todo{lemma for lift-lower}

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen at any point in a process' recursive definition, and they are closed under reflexivity, symmetry and transitivity as shown in Figure \ref{fig:struct-cong}.
In \S \ref{subject-congruence} we will prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then $Q$ is well-typed.
Specifically, in the case of transitivity we must prove that if $P$ is structurally congruent with $Q$ and $Q$ with $R$, and $P$ is well-typed, then so is $R$.
To do so, we will have to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show that the doubly recursive call terminates we index the equivalence relation $=$ by the type $\Rec$, which models the structure of the recursion.

@@

\begin{nilemma}
\todo{introduce entails-unused}
\end{nilemma}

@@

\begin{nitheorem}
  Let $\opctx{\Gamma_l}{\Delta}{\Xi_l}$ and $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  If $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$ then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

@@

Order preserving embeddings model a series of insertions. We only ever need one insertion to prove subject congruence, but there is no loss of generality.

\begin{nitheorem}
  If $\types{\gamma}{\Gamma}{P}{\Xi}$ then $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}

@@

\begin{nitheorem}
  If $\types{\gamma}{\Gamma}{P}{\Xi}$ and $\Unused \; i \; P$, then $\types{delete_i \; \gamma}{delete_i \; \Gamma}{lower_i \; P}{delete_i \; \Xi}$.
\end{nitheorem}

\begin{remark}
  The proof that index $i$ is unused in $P$ must be syntactical and cannot rely on the usage annotation of index $i$ being preserved between $\Gamma$ and $\Xi$: process $P$ can send variable reference $i$ with multiplicity $\zero$ over a channel.
\end{remark}

@@

\begin{proof}
  By induction on \textsc{EQUALS}.
  The case $\constr{struct}$ and its symmetric one under $\constr{sym}$ are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}.
  The case $\constr{comp-assoc}$ is trivial because the encoding with leftover typing of the typing rules naturally preserves associativity.
  The $\constr{comp-sym}$ case for a process $\comp{P}{Q}$ is however more complicated: $P$ must use framing (\S \ref{framing}) to shift its output context to $Q$'s; $Q$ must use framing to shift its input context to $P$'s.
  The middle-point of the resulting derivation on $\comp{Q}{P}$ will no longer be the same.
  The case $\constr{comp-end}$ is trivial because in the typing rule for process $\PO$ the input and output contexts are one and the same.
  In the $\constr{scope-end}$ case, the typing rules $\constr{chan}$ and $\constr{end}$ work together to constraint the multiplicity of the newly create channel to $\ell_{\o}$, making the proof trivial.
  In the opposite direction, we will need to instantiate the newly created channel to a concrete type and multiplicities.
  For this, we will use type $\unit$ and multiplicities $\ell_{\o}$.
  In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we will make use of Lemma $\ref{types-unused}$ to show that as the typing derivation on $P$ will preserved the multiplicity of the recently introduced variable unchanged.
  We will then use strengthening (\S \ref{strengthening}) to delete the recently introduced variable from $P$, and then compose with the typing derivation of $Q$ under $\constr{chan}$.
  In the reverse direction, we will weaken $P$ so that it can accept an additional variable.
  We will then use Lemma \ref{lift-lower} to show that lifting a lowered $P$ leaves $P$ unchanged.
  Finally, for the $\constr{scope-comm}$ rule, we will use swapping as shown in \S \ref{swapping}.
  In the reverse direction, we will use swapping and then show that swapping indices $i$ and $\suc i$ in process $P$ two times leaves $P$ unchanged.
\end{proof}
\todo{make sure constructor names in code coincide}

>>>>

In the \lambdacalc{} $\beta$-reduction operates on syntactically adjacent terms.
In the \picalc{} however, the syntax introduces unnecessary distinctions
(e.g. semantically, parallel composition is defined modulo associativity and commutativity).
There are several ways around this, a structural congruence relation being one of the historical ones.
(Others include labeled transition systems and higher inductive types.)

@@

Structural congruence is a congruent equivalence relation on processes.
Any two structurally congruent processes are strongly bisimilar: they are can follow each other's reduction steps \cite{}.
Figure \ref{fig:struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_i \; Q$ is an inductive proof on $Q$ that witnesses that the index $i$ does not appear nor in the inputs nor in the outputs of the process.
The function $lower \; i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
Finally, $swap \; i \; P$ traverses $P$ (scoped under $\suc \suc n$) and swaps variable references $i$ and $\suc i$.

@@

\begin{nilemma}
  \label{lm:lower-lift}
  The function $lower \; i \; P \; uP$ has an inverse $lift \; i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $lift \; i \; (lower \; i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:swap-swap}
  The function $swap \; i \; P$ is its own inverse: $swap \; i \; (swap \; i \; P) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen at any point in a process' recursive definition, and they are closed under reflexivity, symmetry and transitivity as shown in Figure \ref{fig:struct-cong}.
In \S \ref{subject-congruence} we will prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then $Q$ is well-typed.
Specifically, in the case of transitivity, we must prove that if $P$ is structurally congruent with $Q$ and $Q$ with $R$, and $P$ is well-typed, then so is $R$.
To do so, we will have to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show that the doubly recursive call terminates we index the equivalence relation $=$ by the type $\Rec$, which models the structure of the recursion.

@@

\begin{nilemma}
  \label{lm:types-unused}
  If $P$ is well-typed under input context $\Gamma$ and output context $\Xi$, and variable $i$ is unused within $P$, then $\Gamma_i \equiv \Xi_i$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:types-op}
  For any well typed-process $P$ with input context $\Gamma$ and output context $\Xi$, there exists some $\Delta$ such that $\opctx{\Gamma}{\Delta}{\Xi}$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

Framing captures the idea that the well-typedness of a process does not depend on its unused resources.
These unused resources are not introduced as new variables -- the typing context remains unaltered -- but rather as extra multiplicities that are carried from the input context to the output context.

\begin{nitheorem}
  Let $P$ be a well-typed process from context $\Gamma_l$ to $\Xi_l$.
  Then, per Lemma \ref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

Weakening states that adding new variables to the typing and usage contexts preserves the well-typedness of a process.
The multiplicities of these new variables will be carried from the input context to the output context.
Usually, weakening is presented through order preserving embeddings.
Order preserving embeddings model a series of insertions into a context.
In this paper, we use weakening to prove subject congruence.
To do so, we need of a single insertion into a context, and thus we will limit our theorem statement to that, but there is no loss of generality.

\begin{nitheorem}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let us insert an arbitrary type $t$ at position $i$ in the typing context and an arbitrary multiplicity $x$ at position $i$ into both the input and output contexts.
  Let us reflect this on the process $P$ by incrementing every index greater than or equal to $i$.
  The resulting process will be well-typed so that $\types{insert_i \; t \; \gamma}{insert_i \; x \; \Gamma}{lift_i \; P}{insert_i \; x \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

Strengthening states that removing unused variables from the typing and usage contexts preserves the well-typedness of a process.
The proof that a variable is unused within a process must be syntactical and cannot rely on the preservation of its usage annotations between the input and output contexts: a process can refer to a variable and still make no use of it -- e.g. by sending none of its multiplicities over a channel.

\begin{nitheorem}
  Let $P$ be a well-typed process such that $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable index that is not mentioned by $P$, such that $\Unused \; i \; P$.
  Then removing index $i$ from the typing context and both of the usage contexts and decrementing every variable in $P$ greater than $i$ preserves its well typedness, so that $\types{delete_i \; \gamma}{delete_i \; \Gamma}{lower_i \; P}{delete_i \; \Xi}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{proof}
  By induction on \textsc{EQUALS}.
  The case $\constr{struct}$ and its symmetric one under $\constr{sym}$ are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}.
  The case $\constr{comp-assoc}$ is trivial because the encoding with leftover typing of the typing rules naturally preserves associativity.
  The $\constr{comp-sym}$ case for a process $\comp{P}{Q}$ is however more complicated: $P$ must use framing (\S \ref{framing}) to shift its output context to $Q$'s; $Q$ must use framing to shift its input context to $P$'s.
  The middle-point of the resulting derivation on $\comp{Q}{P}$ will no longer be the same.
  The case $\constr{comp-end}$ is trivial because in the typing rule for process $\PO$ the input and output contexts are one and the same.
  In the $\constr{scope-end}$ case, the typing rules $\constr{chan}$ and $\constr{end}$ work together to constraint the multiplicity of the newly create channel to $\ell_{\o}$, making the proof trivial.
  In the opposite direction, we will need to instantiate the newly created channel to a concrete type and multiplicities.
  For this, we will use type $\unit$ and multiplicities $\ell_{\o}$.
  In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we will make use of Lemma $\ref{lm:types-unused}$ to show that as the typing derivation on $P$ will preserved the multiplicity of the recently introduced variable unchanged.
  We will then use strengthening (\S \ref{strengthening}) to delete the recently introduced variable from $P$, and then compose with the typing derivation of $Q$ under $\constr{chan}$.
  In the reverse direction, we will weaken $P$ so that it can accept an additional variable.
  We will then use Lemma \ref{lm:lower-lift} to show that lifting a lowered $P$ leaves $P$ unchanged.
  Finally, for the $\constr{scope-comm}$ rule, we will use swapping as shown in \S \ref{swapping}.
  In the reverse direction, we will use swapping and then use Lemma \ref{lm:swap-swap} to show that swapping indices twice leaves $P$ unchanged.
\end{proof}
\todo{make sure constructor names in code coincide}

\begin{remark}
  We encode the recursive structure of \textsc{Equals} in its inductive definition -- by indexing its type with \textsc{Rec}.
  This acts as a justification of the well-formedness of the recursion scheme, which Agda cannot immediately see as terminating due to the interaction between the transitive and symmetric cases.
\end{remark}
